Intrusion Detection with \\ Genetic Algorithms and Fuzzy Logic

### 2013-11-23 paper.tex
Added some stuff to the results section.

\section{Introduction}
 \mycomment{WORK ON INTRODUCTION} The CSci computer lab at the University of Minnesota, Morris gets large numbers of login attempts that are attempts at intrusion. An attacker could be trying to gain root access to the system, which would then give them the ability to do things like delete files of other users and change user passwords. If the university had an intrusion detection system, it would be possible to classify those attempts into legitimate and illegitimate attempts to login. Then it would be possible to block IP addresses that are generating large numbers of attacks.

@@

In the research discussed in Section~\ref{sec:fuzGenAlgImp}~\cite{6496342, 6559603}, eight of the 41 KDD99 features were used based on research done in~\cite{Fries:2008:FAN:1388969.1389037}. These eight features are:
\begin{enumerate}
  \item duration: the length of the record in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts in this record.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections in other records.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure to get an ACK back.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service. \mycomment{Have example of service}
\end{enumerate}

@@

\begin{table*}
\center
\caption{Results from Experiment 1, using only RLD09~\cite{6496342}}
\vspace{0.20cm}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

@@

\begin{table}
\caption{Unknown Attack Experiment, using only RLD09~\cite{6496342}}
\vspace{0.20cm}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

7 & Http Flood (DoS) & Avg =          & Avg =\\
  & RCP Scan (Probe)  & 99.94 & 97.50\\
  & Fin Scan (Probe) &                &\\
\hline\end{tabular}
\label{tab:fuzGenExp2}
\end{table}

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks~\cite{6559603}}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
3 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
4 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\end{tabular}
\label{tab:kddAttacks}
\end{table}

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm to classify types of attacks. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of a majority of the cases were greater than 97.5\%. Again, there were only two test cases that had low detection rates, one of which is case 2 in Table~\ref{tab:rldAttacks}. The two low detection rates were 89.59\% (denial of service), and 86.89\% (probe).

\begin{table}
\caption{Results for RLD09 with Certain Attacks~\cite{6559603}}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Smurf & DoS & 0.02 & 0 & 99.98\\
2 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
3 & Ackscan & Probe & 0.03 & 0 & 99.97\\
4 & Synscan & Probe & 0.65 & 4.2 & 99.24\\
\end{tabular}
\label{tab:rldAttacks}
\end{table}

@@

\subsection{Algorithm Overview}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used a genetic algorithm to develop an IDS.

@@

\begin{table}
\center
\caption{Training and testing records used in~\cite{DBLP:journals/corr/abs-1204-1336}}
\vspace{0.20cm}
\begin{tabular}{lll}
  & Training & Testing \\ 
Normal & 97,280 & 60,593\\ \hline
DoS	   & 391,458 & 229,853\\
R2L    & 1,124   & 16,189\\
U2R    & 52      & 228\\
Probe  & 4,107  & 4,166\\ \hline
Total Attacks & 396,741 & 250,436 \\
Total Records  & 494,021 & 311,029\\
\end{tabular}
\center
\label{tab:numberOfRecords}
\end{table}

@@

\begin{table*}
\center
\caption{Results for GA Experiment in~\cite{DBLP:journals/corr/abs-1204-1336}}
\vspace{0.20cm}
\begin{tabular}{cc|cccccc}
& & & & Predicted & & & \\
& Actual & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{3-8}
& Normal & 42138 & 1421 & 15835 & 486 & 713 & 69.5\\
& Probe & 398 & 2963 & 654 & 2 & 149 & 71.1   \\
& Dos & 921 & 432 & 228489 & 1 & 10 & 99.4\\
& U2R & 146 & 21 & 8 & 43 & 10 & 18.9\\
& R2L & 11191 & 578 & 3398 & 141 & 881 & 5.4\\
&& &&&&& \\
& \% Correct &  76.9 & 54.7 & 92.0 & 6.4 & 50.0 & \\
\end{tabular}
\center
\label{tab:genAlgResults}
\end{table*}

@@

\begin{table}
\center
\caption{Accuracy of the GA in~\cite{DBLP:journals/corr/abs-1204-1336}}
\vspace{0.20cm}
\begin{tabular}{l|ll}
& & Predicted\\
Actual & Normal & Attack \\ \cline{2-3}
Normal & TN: 42,138. 69.5\% & FP: 18,455. 30.5\% \\
Attack & FN: 12,528. 5\% & TP: 237,908. 94.9\%  \\
\end{tabular}
\center
\label{tab:genAlgResults2}
\end{table}

The authors of~\cite{DBLP:journals/corr/abs-1204-1336} compared their results (Table~\ref{tab:genAlgResults}) with the winning entry of the KDD99 Classifier Learning Contest~\cite{KDD99Contest}, which can be seen in Table~\ref{tab:winningEntry}. The winning entry used a decision tree algorithm. The winning entry had a higher detection rate for normal activity, probe attacks, and remote to user attacks. The authors of~\cite{DBLP:journals/corr/abs-1204-1336} found that they had a better detection rate for denial of service and user to root attacks than the winning entry. For the winning entry, the detection rate of denial of service was 97.1\%, and for user to root it was 13.2\%. In~\cite{DBLP:journals/corr/abs-1204-1336} the detection rate of denial of service was 99.4\% and for user to root it was 18.9\%. The authors of~\cite{DBLP:journals/corr/abs-1204-1336} provide no information about the statistical significance of these improvements.

\begin{table*}
\center
\caption{Results for the Winning Entry of the KDD99 Classifier Learning Contest}
\vspace{0.20cm}
\begin{tabular}{c|cccccc}
& & & & Predicted & & \\
Actual & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{2-7}
Normal &  60262 & 243 & 78 & 4 & 6 & 99.5\\
Probe &  511 & 3471 & 184 & 0 & 0 & 83.3 \\
Dos &  5299 & 1328 & 223226 & 0 & 0 & 97.1\\
U2R &  168 & 20 & 0 & 30 & 10 & 13.2\\
R2L &  14527 & 294 & 0 & 8 & 1360 & 8.4\\
& &&&&& \\
\% Correct &  74.6 & 64.8 & 99.9 & 71.4 & 98.8 &\\
\end{tabular}
\center
\label{tab:winningEntry}
\end{table*}

@@

>>>>

\section{Introduction}
 \mycomment{work on introduction} The CSci computer lab at the University of Minnesota, Morris gets large numbers of login attempts that are attempts at intrusion. An attacker could be trying to gain root access to the system, which would then give them the ability to do things like delete files of other users and change user passwords. If the university had an intrusion detection system, it would be possible to classify those attempts into legitimate and illegitimate attempts to login. Then it would be possible to block IP addresses that are generating large numbers of attacks.

@@

In the research discussed in Section~\ref{sec:fuzGenAlgImp}~\cite{6496342, 6559603}, eight of the 41 KDD99 features were used based on research done in~\cite{Fries:2008:FAN:1388969.1389037}. These eight features are:
\begin{enumerate}
  \item duration: the length of the record in seconds.
  \item src\_bytes: the number of bytes sent from source to destination. Source is the user who may or may not be an attacker, and destination is the server being potentially attacked.
  \item num\_failed\_logins: the number of failed login attempts in this record.
  \item root\_shell: returns 1 if root shell is obtained, which means that the user is able to log in as root. If root shell is not obtained, 0 is returned.
  \item num\_access\_files: the number of operations on access control files. Access control files specify which users are granted access to objects and what operations are allowed on the objects. An example would be (Alice, delete), which would give Alice permission to delete the file.~\cite{accessControl}
  \item srv\_count: the number of connections in other records.
  \item serror\_rate: the percentage of connections that have ``SYN" errors. When a client attempts to connect to a server, it first sends a SYN (synchronize) message to the server. The server then acknowledges the request by sending a SYN-ACK to the client. The connection is established when the client sends an ACK back to the server. A SYN error is a failure to get an ACK back.~\cite{TCP}
  \item same\_srv\_rate: the percentage of connections to the same service.
\end{enumerate}

@@

\begin{table*}
\center
\caption{Results from Experiment 1, using only RLD09~\cite{6496342}.}
\vspace{0.20cm}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

@@

\begin{table}
\caption{Unknown Attack Experiment, using only RLD09~\cite{6496342}.}
\vspace{0.20cm}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

7 & HTTP Flood (DoS) & Avg =          & Avg =\\
  & RCP Scan (Probe)  & 99.94 & 97.50\\
  & Fin Scan (Probe) &                &\\
\hline\end{tabular}
\label{tab:fuzGenExp2}
\end{table}

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks. 10 tests were run in total, 5 are shown here.~\cite{6559603}}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
3 & Neptune & DoS & 0.15 & 0.34 & 99.75\\
4 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
5 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\end{tabular}
\label{tab:kddAttacks}
\end{table}

The final experiment that was run used only the RLD09 data set with the fuzzy genetic algorithm to classify types of attacks. 17 tests were run, and Table~\ref{tab:rldAttacks} shows the accuracy of detecting some of the cases. The results showed that the detection rate of 15 out of 17 cases were greater than 97\%. Again, there were only two test cases that had low detection rates, cases 3 and 5 in Table~\ref{tab:rldAttacks}. The two low detection rates were 89.59\% (denial of service), and 86.89\% (probe).

\begin{table}
\caption{Results for RLD09 with Certain Attacks. 17 tests were run in total, 6 are shown here.~\cite{6559603}}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & HTTP Flood & DoS & 0.36 & 3.5 & 99.46\\
2 & Smurf & DoS & 0.02 & 0 & 99.98\\
3 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
4 & Fin Scan & Probe & 2.58 & 0 & 97.50\\
5 & IP Scan & Probe & 13.01 & 16.4 & 86.89\\
6 & Syn Scan & Probe & 0.65 & 4.2 & 99.24\\
\end{tabular}
\label{tab:rldAttacks}
\end{table}

@@

The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used a genetic algorithm to develop an IDS. They didn't explain their research that well and left a lot of things out, so I will just be describing their results.

@@

\begin{table}
\center
\caption{Training and testing records used in~\cite{DBLP:journals/corr/abs-1204-1336}.}
\vspace{0.20cm}
\begin{tabular}{lll}
  & Training & Testing \\ 
Normal & 97,280 & 60,593\\ \hline
DoS	   & 391,458 & 229,853\\
R2L    & 1,124   & 16,189\\
U2R    & 52      & 228\\
Probe  & 4,107  & 4,166\\ \hline
Total Attacks & 396,741 & 250,436 \\
Total Records  & 494,021 & 311,029\\
\end{tabular}
\center
\label{tab:numberOfRecords}
\end{table}

@@

\begin{table*}
\center
\caption{Results for GA Experiment in~\cite{DBLP:journals/corr/abs-1204-1336}.}
\vspace{0.20cm}
\begin{tabular}{cc|cccccc}
& & & & Predicted & & & \\
& Actual & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{3-8}
& Normal & 42138 & 1421 & 15835 & 486 & 713 & 69.5\\
& Probe & 398 & 2963 & 654 & 2 & 149 & 71.1   \\
& Dos & 921 & 432 & 228489 & 1 & 10 & 99.4\\
& U2R & 146 & 21 & 8 & 43 & 10 & 18.9\\
& R2L & 11191 & 578 & 3398 & 141 & 881 & 5.4\\
&& &&&&& \\
& \% Correct &  76.9 & 54.7 & 92.0 & 6.4 & 50.0 & \\
\end{tabular}
\center
\label{tab:genAlgResults}
\end{table*}

@@

\begin{table}
\center
\caption{Accuracy of the GA in~\cite{DBLP:journals/corr/abs-1204-1336}.}
\vspace{0.20cm}
\begin{tabular}{l|ll}
& & Predicted\\
Actual & Normal & Attack \\ \cline{2-3}
Normal & TN: 42,138. 69.5\% & FP: 18,455. 30.5\% \\
Attack & FN: 12,528. 5\% & TP: 237,908. 94.9\%  \\
\end{tabular}
\center
\label{tab:genAlgResults2}
\end{table}

The authors of~\cite{DBLP:journals/corr/abs-1204-1336} compared their results (Table~\ref{tab:genAlgResults}) with the winning entry of the KDD99 Classifier Learning Contest~\cite{KDD99Contest}, which can be seen in Table~\ref{tab:winningEntry}. The winning entry used a decision tree algorithm. The winning entry had a higher detection rate for normal activity, probe attacks, and remote to user attacks. The authors of~\cite{DBLP:journals/corr/abs-1204-1336} found that they had a better detection rate for denial of service and user to root attacks than the winning entry. For the winning entry, the detection rate of denial of service was 97.1\%, and for user to root it was 13.2\%. In~\cite{DBLP:journals/corr/abs-1204-1336} the detection rate of denial of service was 99.4\% and for user to root it was 18.9\%. The authors of~\cite{DBLP:journals/corr/abs-1204-1336} provide no information about the statistical significance of these improvements.
\mycomment{compare~\cite{DBLP:journals/corr/abs-1204-1336} with~\cite{6496342, 6559603}}

\begin{table*}
\center
\caption{Results for the Winning Entry of the KDD99 Classifier Learning Contest.}
\vspace{0.20cm}
\begin{tabular}{c|cccccc}
& & & & Predicted & & \\
Actual & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{2-7}
Normal &  60262 & 243 & 78 & 4 & 6 & 99.5\\
Probe &  511 & 3471 & 184 & 0 & 0 & 83.3 \\
Dos &  5299 & 1328 & 223226 & 0 & 0 & 97.1\\
U2R &  168 & 20 & 0 & 30 & 10 & 13.2\\
R2L &  14527 & 294 & 0 & 8 & 1360 & 8.4\\
& &&&&& \\
\% Correct &  74.6 & 64.8 & 99.9 & 71.4 & 98.8 &\\
\end{tabular}
\center
\label{tab:winningEntry}
\end{table*}

@@

\mycomment{fix references}
